A thermodynamic view of dusty protoplanetary disks

### 2017-08-07 model.tex
last head to toe edit before submitting, in progress

The dust and gas fluids interact via a drag force parameterized by the relative stopping 
time $\tstop$ such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\rho}\frac{\left(\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g}\right)}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume.%, where
%\begin{align}
%  \widetilde{\bm{v}} \equiv\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} 
%\end{align}
%is the dust-gas velocity difference. 
 Note that $\tstop$ 
differs slightly from the \emph{particle} stopping time $\tau_\mathrm{s} =
\tstop\rho/\rhog$ used in some studies
\citep[e.g.][]{youdin05a}.

In general the relative velocity $\bm{v}_\mathrm{d} -
\bm{v}_\mathrm{g}$ obeys a complicated evolutionary equation 
\citep[see, e.g.][]{youdin05a}. % obtained from taking the difference
%between evolutionary equations for $\bm{v}_\mathrm{g}$ and
%$\bm{v}_\mathrm{d}$. 
 However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency, we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{jacquet11}. This equation reflects the well-known effect of 
particle drift towards pressure maxima \citep{weidenschilling77}.

@@

Eq. \ref{dusteq} results from mass conservation of dust, where 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon} 
\end{align}
is the dust-fraction and $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. If $\tstop=0$ then the 
dust-to-gas ratio is conserved following the fluid. Otherwise, 
$\epsilon$ evolves due to particle drift in response to pressure
gradients. Note that Eq. \ref{dusteq} is equivalent to Eq. 46 of 
\cite{jacquet11}.

The total gravitational potential is $\Phi_\mathrm{tot}=\Phi +
\psi$. Here $\Phi$ originates from a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right),  
\end{align}
where $(r,\phi, z)$ are cylindrical co-ordinates centered on the star, and 
$G$ is the gravitational constant. The second equality is the  
thin-disk approximation, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria (\S\ref{steady_state}). 
The disk's own gravitational potential $\psi$ is included for
completeness, but we will mostly neglect it unless stated
otherwise. In such cases $\psi$ satisfies the Poisson equation 
\begin{align}
  \nabla^2\psi = 4 \pi G \rho.\label{poisson}
\end{align}

For the mixture's temperature evolution, Eq. \ref{tempeq}, $\gamma$ is
the adiabatic index; $\mathcal{H}$ represents heating (which would
include that due to dust-gas friction) and $\mathcal{H}_\mathrm{eff}$
is an effective source term arising from transforming from the
two-fluid to one-fluid formalism \citep[see][ for
details]{laibe14}. We also include a simple model of radiative
cooling, 
\begin{align}
  \Lambda = \frac{T -
    T_\mathrm{ref}}{t_\mathrm{cool}}, \label{cooling} 
\end{align}
which relaxes the mixture back to a prescribed temperature profile   
$T_\mathrm{ref}$ on a timescale of $\tcool$. We will shortly simplify
the problem by considering rapid cooling, $\tcool\to0$.

Eq. \ref{masseq}---\ref{tempeq} are not yet equivalent to standard 
hydrodynamics, which typically evolves two scalar fields: the density and
temperature (or pressure). By contrast, Eq. \ref{masseq}---\ref{tempeq}
involves 3 scalars: $\rho$, $\tepsilon$, and $T$. 
However, we can establish an  analogy by fixing the gas
equation of state, thus eliminating Eq. \ref{tempeq}; but then reform 
the dust-to-gas ratio evolution equation (\ref{dusteq}) into an
energy-like equation.

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
We simplify the gas energy equation by considering short cooling 
times $\tcool\to 0$, appropriate for the outer parts of an irradiated
protoplanetary disk \citep{chiang97,lin15}. Then the disk temperature
$T = T_\mathrm{ref}$ at all times, and so we may
adopt a locally isothermal equation of state 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}\left(1 - \tepsilon\right)\rho,   
\end{align}
where $c_s(r,z)= \sqrt{\mathcal{R}T_\mathrm{ref}/\mu}$ is a prescribed
sound-speed profile fixed in time. In most applications we consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

@@

\subsection{Effective energy equation}\label{energy_analogy}
Although we have deleted the true energy
equation (Eq. \ref{tempeq}) by fixing a locally isothermal
equation of state (Eq. \ref{eos}), we show that the mixture 
nevertheless obeys an effective energy equation. This is
because advection of the dust-fraction, described by
Eq. \ref{dusteq}, can be transformed into an energy-like 
equation. 
%We show that for a prescribed temperature distribution, the mixture
%obeys an evolutionary energy equation, due to the advection of the
%dust-fraction. 
Eliminating $\tepsilon$ in favor of $P$, 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
%\frac{DP}{Dt} 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  
&= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align} 
For comparison with the standard energy equation in hydrodynamics,
re-writing Eq. \ref{tempeq} for a pure ideal gas with $P \propto
\rhog T$, and reverting back to gas velocities, gives  
%For later comparison it is convenient to re-write the temperature
%evolution, Eq. \ref{tempeq}, to that for the pressure: 
\begin{align}
  \frac{\p P}{\p t} + \bm{v}_\mathrm{g}\cdot\nabla P = - \gamma P
  \nabla\cdot \bm{v}_\mathrm{g}  +
  \frac{P}{T}\left(\mathcal{H} -
    \Lambda\right).\label{real_pressure} 
%\frac{\rho \mathcal{R}}{\mu}\frac{T
%  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align} 
We can thus interpret Eq. \ref{eff_energy} as the energy equation for
and ideal gas with adiabatic index $\gamma=1$, but now with the
imposed temperature gradient $\nabla c_s^2$ and dust-gas drag $\mathcal{C}$ 
acting as source terms.

>>>>

The dust and gas fluids interact via a drag force parameterized by the relative stopping 
time $\tstop$ such that 
\begin{align}  
  \rhod\left.\frac{\p \bm{v}_\mathrm{d}}{\p t}\right|_\mathrm{drag}= -
  \rhog\left.\frac{\p \bm{v}_\mathrm{g}}{\p t}\right|_\mathrm{drag}=
  - \frac{\rhog\rhod}{\rho}\frac{\left(\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g}\right)}{\tstop}. 
\end{align}
is the dust-gas friction force per unit volume. %, where
%\begin{align}
%  \widetilde{\bm{v}} \equiv\bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} 
%\end{align}
%is the dust-gas velocity difference. 
 Note that $\tstop$ 
differs slightly from the \emph{particle} stopping time $\tau_\mathrm{s} =
\tstop\rho/\rhog$ used in some studies
\citep[e.g.][]{youdin05a}.

In general the relative velocity $\bm{v}_\mathrm{d} -
\bm{v}_\mathrm{g}$ obeys a complicated evolutionary equation 
\citep[see, e.g.][]{youdin05a}. % obtained from taking the difference
%between evolutionary equations for $\bm{v}_\mathrm{g}$ and
%$\bm{v}_\mathrm{d}$. 
 However, for tightly-coupled dust
particles with $\tstop\OmK\ll 1 $, where $\OmK$ is the Keplerian orbital
frequency (since we are interested in protoplanetary disks), we can use the 
`terminal velocity approximation' to set %The dust-gas velocity difference 
%Then so that the dust-gas velocity
%difference is 
\begin{align}
  \bm{v}_\mathrm{d} - \bm{v}_\mathrm{g} = \frac{\nabla
    P}{\rhog}\tstop 
\end{align}
\citep{jacquet11}. This equation reflects the well-known effect of 
particle drift towards pressure maxima \citep{weidenschilling77}.

@@

Eq. \ref{dusteq} is obtained from the dust continuity equation,  
$\p_t\rhod + \nabla\cdot\left(\rhod\bm{v}_\mathrm{d}\right)= 0$, 
by writing 
$\bm{v}_\mathrm{d}=\bm{v} + \tstop\nabla P/\rho$, and eliminating  
$\rhod$ in favor of the dust fraction $\tepsilon$: 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon},
\end{align}
where $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. 
If $\tstop=0$ then the 
dust-to-gas ratio is conserved following the fluid. Otherwise, 
$\epsilon$ evolves due to particle drift in response to pressure
gradients. Note that Eq. \ref{dusteq} is equivalent to Eq. 46 of 
\cite{jacquet11}.

The total gravitational potential $\Phi_\mathrm{tot}=\Phi +
\psi$ includes that from a central star of mass $M_*$ and the disk's own
potential $\psi$. We adopt 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right),  
\end{align}
where $(r,\phi, z)$ are cylindrical co-ordinates centered on the star, and 
$G$ is the gravitational constant. The second equality is the  
thin-disk approximation, appropriate for
$|z|\ll r$. We use this approximate form in order to
obtain explicit expressions for disk equilibria (\S\ref{steady_state}). 
The disk potential $\psi$ satisfies the Poisson equation
\begin{align}
  \nabla^2\psi = 4 \pi G \rho.\label{poisson}
\end{align}
We include $\psi$ for
completeness, but we will mostly neglect it unless stated
otherwise.

For the mixture's temperature evolution, Eq. \ref{tempeq}, reflects the gas energy equation. 
Here $\gamma$ is 
the adiabatic index; $\mathcal{H}$ represents gas heating and $\mathcal{H}_\mathrm{eff}$ 
is an effective source term arising from transforming from the 
two-fluid to one-fluid variables \citep[see][ for 
details]{laibe14}. We also include a simple model of radiative
cooling, 
\begin{align}
  \Lambda = \frac{T -
    T_\mathrm{ref}}{t_\mathrm{cool}}, \label{cooling} 
\end{align}
which relaxes the mixture back to a prescribed temperature profile   
$T_\mathrm{ref}$ on a timescale of $\tcool$. We will shortly simplify
the problem by considering rapid cooling, $\tcool\to0$.

Eq. \ref{masseq}---\ref{tempeq} are not yet equivalent to standard 
hydrodynamics, which typically evolves two scalar fields: the density and
temperature (or pressure). By contrast, Eq. \ref{masseq}---\ref{tempeq}
involves three scalars: $\rho$, $\tepsilon$, and $T$. 
However, we can establish an  analogy by fixing the gas
equation of state, thus eliminating Eq. \ref{tempeq}; but then reform 
the dust-to-gas ratio evolution equation (\ref{dusteq}) into an
energy-like equation.

\subsection{Locally isothermal equation of state}\label{loc_iso_eos}
We simplify the energy equation by considering short cooling 
times $\tcool\to 0$, appropriate for the outer parts of an irradiated
protoplanetary disk \citep{chiang97,lin15}. Then the disk temperature
$T = T_\mathrm{ref}$ at all times, and so we may
adopt a locally isothermal equation of state 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}\left(1 - \tepsilon\right)\rho,   
\end{align}
where $c_s(r,z)= \sqrt{\mathcal{R}T_\mathrm{ref}/\mu}$ is a prescribed
sound-speed profile fixed in time. In most applications we consider vertically 
isothermal disks with \begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature. For $q=0$
the disk is strictly isothermal.

@@

\subsection{Effective energy equation}\label{energy_analogy}
Although we have deleted the true energy
equation (Eq. \ref{tempeq}) by fixing a locally isothermal
equation of state (Eq. \ref{eos}), we show that the mixture 
nevertheless obeys an effective energy equation. This is
because advection of the dust-fraction, described by
Eq. \ref{dusteq}, can be transformed into an energy-like 
equation. 
%We show that for a prescribed temperature distribution, the mixture
%obeys an evolutionary energy equation, due to the advection of the
%dust-fraction. 
Writing $\tepsilon=\tepsilon(P,\rho)$ using Eq. \ref{eos}
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
%\frac{DP}{Dt} 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  
&= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align} 
For comparison with the standard energy equation in hydrodynamics,
re-writing Eq. \ref{tempeq} for a pure ideal gas with $P \propto
\rhog T$, and reverting back to gas velocities, gives  
%For later comparison it is convenient to re-write the temperature
%evolution, Eq. \ref{tempeq}, to that for the pressure: 
\begin{align}
  \frac{\p P}{\p t} + \bm{v}_\mathrm{g}\cdot\nabla P = - \gamma P
  \nabla\cdot \bm{v}_\mathrm{g}  +
  \frac{P}{T}\left(\mathcal{H} -
    \Lambda\right).\label{real_pressure} 
%\frac{\rho \mathcal{R}}{\mu}\frac{T
%  - T_\mathrm{ref}}{t_\mathrm{cool}}, \label{realenergy}
\end{align} 
We can thus interpret Eq. \ref{eff_energy} as the energy equation for
and ideal gas with adiabatic index $\gamma=1$, but now with the
imposed temperature gradient $\nabla c_s^2$ and dust-gas drag $\mathcal{C}$ 
acting as source terms.

### 2017-08-07 paper.tex
last head to toe edit before submitting, in progress

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follow the gas flow.
  %AY: Edited b/c "This is" was too vague
  This near conservation of dust-to-gas ratio along streamlines
  is analogous to the near conservation of entropy along flows of (dust-free) gas with weak heating and cooling.   
  We develop this thermodynamic analogy into a
  framework to study dusty gas dynamics in protoplanetary disks. We  
  show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this correspondence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 perfectly coupled, thin dust layers cannot cause axisymmetric 
 instabilties; 
\item 
 radial dust edges are unstable if dust is vertically well-mixed;  
\item 
  the streaming instability neccessarily involves a gas pressure
  response 
  that lags behind dust density; 
%  the streaming instability neccessarily involves a phase lag between
%  gas pressure and dust densities; 
\item dust-loading introduces buoyancy forces that generally 
  stabilizes the vertical shear 
  instability associated with global radial temperature gradients. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss %future applications of this thermodynamic framework in
%finding 
 dusty analogs of other hydrodynamic 
processes (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and how to simulate dusty protoplanetary disks
with minor tweaks to existing codes for pure gas dynamics.%, as well as possible generalizations
%of our analogy between dust-laden and nearly adiabatic flows.

@@

The two-fluid equations for dust and gas can be reformulated into 
equivalent dynamical equations for the center of mass velocity $\bm{v}$
and the relative dust-gas velocity  $\bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}$ 
\citep{youdin05a}. In this reformulation, continuity equations for dust and gas can be replaced 
by the mass conservation of total density $\rho$ plus the evolution of the dust-to-gas ratio 
(or dust fraction, \citealp{laibe14}).

@@

The purpose of our physical and mathematical parallel between
dusty gas and standard hydrodynamics is to gain a deeper understanding of the dynamics
and stability of dusty disks.  
We can apply established methods to generalize previous stability 
results for pure gas dynamics to dusty disks.  
We show that strictly isothermal protoplanetary disks with
perfectly-coupled dust are generally 
stable to axisymmetric perturbations, unless the 
dust-to-gas ratio varies more rapidly in radius than in height.  
%, regardless of how  thin the dust
%layer is. 
%{\bf should we mention vertically well mixed case here?} 
We put forward a thermodynamical interpretation of 
dust-gas drag instabilities, and apply it to the streaming instability.   
We also study the effect of dust-loading on the vertical shear
instability previously considered in pure gas disks
\citep[VSI,][]{nelson13,lin15,barker15}.

@@

There are several situations where the pressure and density of a fluid
are not in phase. The obvious case is if the fluid is subject to
external heating/cooling. For example, in strongly irradiated
protoplanetary disks the
disk temperature  $T(r)$ is time-independent but varies with the cylindrical radius
$r$ from the star \citep{chiang97}. Since $P\propto \rhog T$ for an
ideal gas, %the pressure of a gas parcel varies with  
%position as well. 
there is no reason to expect $P$ and $\rhog$ to 
be in phase as a gas parcel oscillates between different radii. In
fact, we will show that this is a fundamental property of the vertical
shear instability \citep{lin15}.

@@

Another possibility, as annotated in Fig. \ref{pdv_cartoon}, is a dusty gas. The
relevant density here is the total density $\rho = \rhod + \rhog$, 
but the fluid pressure $P$ is due to gas only. If dust were 
perfectly-coupled to the gas, then $P$ and $\rho$ would be in phase, 
and there is no work done. However, for finite dust-gas drag, $\rhod$ and
$\rhog$ are not neccessarily in phase, because dust particles can
drift relative to the gas.   
\emph{If} this causes $P$ to lag 
$\rho$, then the positive work done would lead to growing
oscillations. Indeed, we show that this is true for the streaming
instability.

In order to apply this thermodynamic interpretation of dust-gas drag
instabilities, we need to develop a formal analogy between dusty-gas
and pure gas dynamics. We show this is possible in the limit of
strong drag and a fixed gas equation of state.

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follow the gas flow.
  %AY: Edited b/c "This is" was too vague
  This near conservation of dust-to-gas ratio along streamlines
  is analogous to the near conservation of entropy along flows of (dust-free) gas with weak heating and cooling.   
  We develop this thermodynamic analogy into a
  framework to study dusty gas dynamics in protoplanetary disks. We  
  show that an isothermal dusty gas behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this correspondence 
  to deduce that   
\begin{inparaenum}[1)] 
\item 
 perfectly coupled, thin dust layers cannot cause axisymmetric 
 instabilties; 
\item 
 radial dust edges are unstable if the dust is vertically well-mixed;  
\item 
  the streaming instability neccessarily involves a gas pressure
  response 
  that lags behind dust density; 
%  the streaming instability neccessarily involves a phase lag between
%  gas pressure and dust densities; 
\item dust-loading introduces buoyancy forces that generally 
  stabilizes the vertical shear 
  instability associated with global radial temperature gradients. 
\end{inparaenum}  
%growth rates, and 
%preferentially stabilize modes with small radial 
%wavelengths, as well as `surface modes' concentrated near the
%disk boundaries. 
%These results can be understood in terms of 
%buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also discuss %future applications of this thermodynamic framework in
%finding 
 dusty analogs of other hydrodynamic 
processes (e.g. Rossby wave instability, convective overstability, and 
zombie vortices), and how to simulate dusty protoplanetary disks
with minor tweaks to existing codes for pure gas dynamics.%, as well as possible generalizations
%of our analogy between dust-laden and nearly adiabatic flows.

@@

The two-fluid equations for dust and gas can be reformulated into 
equivalent dynamical equations for the center of mass velocity $\bm{v}$
and the relative dust-gas velocity  $\bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}$ 
\citep{youdin05a}. In this reformulation, continuity equations for dust and gas can be replaced 
by the mass conservation of total density $\rho$ plus the evolution of the dust-to-gas ratio $\rhod/\rhog$ 
(or dust fraction, $\rhod/\rho$, \citealp{laibe14}).

@@

The purpose of our physical and mathematical parallel between
dusty gas and standard hydrodynamics is to gain a deeper understanding of the dynamics
and stability of dusty protoplanetary disks.  
We can apply established methods to generalize previous stability 
results for pure gas dynamics to dusty disks.  
We show that strictly isothermal disks with 
perfectly-coupled dust are generally 
stable to axisymmetric perturbations, unless the 
dust-to-gas ratio varies more rapidly in radius than in height.  
%, regardless of how  thin the dust
%layer is. 
%{\bf should we mention vertically well mixed case here?} 
We put forward a thermodynamical interpretation of 
dust-gas drag instabilities, and apply it to the streaming instability.   
We also study the effect of dust-loading on the vertical shear
instability previously considered in pure gas disks
\citep[VSI,][]{nelson13,lin15,barker15}.

@@

There are several situations where the pressure and density of a fluid
are not in phase. The obvious case is if the fluid is subject to
external heating/cooling. For example, in strongly irradiated
protoplanetary disks the
disk temperature  $T(r)$ is time-independent but varies with the cylindrical radius
$r$ from the star \citep{chiang97}. Since $P\propto \rhog T$ for an
ideal gas, %the pressure of a gas parcel varies with  
%position as well. 
there is no reason to expect $P$ and $\rhog$ to 
be in phase as a gas parcel oscillates between different radii and adopt the corresponding local temperatures. 
 In 
fact, we will show that this is a fundamental property of the vertical
shear instability \citep{lin15}.

@@

Another possibility, as annotated in Fig. \ref{pdv_cartoon}, is a dusty gas. The
relevant density here is the total density $\rho = \rhod + \rhog$, 
but the fluid pressure $P$ is due to gas only. If dust were 
perfectly-coupled to the gas, then $P$ and $\rho$ would be in phase, 
and there is no work done. However, for finite dust-gas drag, $\rhod$ and
$\rhog$ are not neccessarily in phase, because dust particles can
drift relative to the gas.   
If this causes $P$ to lag 
$\rho$, then the positive work done would lead to growing
oscillations. Indeed, we show that this is true for the streaming
instability.

In order to apply this thermodynamic interpretation of dust-gas drag
instabilities, we need to develop a formal analogy between dusty-gas
and pure hydrodynamics. We show this is possible in the limit of
strong drag and a fixed gas equation of state.
